---
title: "Application Exercise 13 - Data wrangling with tidyr"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

```{r}

```

# Part 1: (Atomic) Vectors

Atomic vectors, of which there are six types: logical, integer, double, character, complex, and raw.

Integer and double vectors are collectively known as numeric vectors.

-   lgl: logical
-   int: integer
-   dbl: double
-   chr: character

```{r}

```

## Types of atomic vectors

```{r}


```

## Explicit vector coercion & augmented vectors

Vectors can also contain arbitrary additional metadata in the form of attributes. These attributes are used to create augmented vectors which build on additional behaviour. For example, factors are built on top of integer vectors.

```{r}


```

# Part 2: tidyr R Package

```{r}
# Data source: https://github.com/rbtl-fs22/research-project-template-disposa-liz/blob/main/data/raw_data/rbtl_fs22_waste_composition_bellevue.csv

waste_data <- read_csv("data/waste-characterisation-disposa-liz.csv")

```

## Example: Waste characterisation data

The data for this exercise is taken from: https://github.com/rbtl-fs22/research-project-template-disposa-liz

```{r}


```

## Example: Gapminder

The following exercise is taken from @naupaka_zimmerman_2019_3265164.

```{r}

gapminder_wide <- read_csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/data/gapminder_wide.csv")

```

```{r}



```

### From wide to long

```{r}



```

**Your turn**

Using `gapminder_long`, calculate the mean life expectancy, population, and gdpPercap for each continent. **Hint:** use the `group_by()` and `summarize()` functions we learned in the `dplyr` lesson.

```{r}



```

### From long to intermediate format

```{r}


```

# Part 3: Working with multiple dataframes

## Example: Waste characterisation data

The data for this exercise is taken from: https://github.com/rbtl-fs22/research-project-template-lord-of-the-bins/

```{r}
waste_data_lord <- read_rds("data/raw_data/waste-characterisation-lord-of-the-bins-tidy.rds")
erz_bins <- read_csv("data/raw_data/erz.abfallgefaess_p.csv")
```


```{r}


```

## Example: Band members

This exercise is copied from: https://dplyr.tidyverse.org/reference/mutate-joins.html#ref-examples

```{r}


```

# Reference
